The Birth of a Galaxy: Primordial Metal Enrichment and Stellar Populations

### 2011-07-06 ApJ/ms.tex
Adding halo enrichment figure and some text on the cosmic SFR.

\author{John H. Wise\altaffilmark{1,2}, 
  Matthew J. Turk\altaffilmark{3},
  Michael L. Norman\altaffilmark{3},
  Tom Abel\altaffilmark{4}}

\altaffiltext{1}{Department of Astrophysical Sciences, Princeton
  University, Peyton Hall, Ivy Lane, Princeton, NJ 08544}
\altaffiltext{2}{Hubble Fellow}
\altaffiltext{3}{Center for Astrophysics and Space Sciences,
  University of California at San Diego, La Jolla, CA 92093}
\altaffiltext{4}{Kavli Institute for Particle Astrophysics and
  Cosmology, Stanford University, Menlo Park, CA 94025}
\email{jwise@astro.princeton.edu}

@@

We use mass-dependent luminosities and lifetimes of the Pop III stars
from \citet{Schaerer02}.  The radiation field is evolved with adaptive
ray tracing \citep{Abel02_RT, Wise10} and is coupled self-consistently
to the hydrodynamics.  We model the \hh~dissociating radiation with an
optically-thin, inverse square profile, centered on all star
particles.  These stars die as pair-instability SNe (PISNe) if they
are in the mass range 140--260 \Ms~\citep{Heger03}.  We use the
explosion energy from \citet{Heger02}, where we fit the following
function to their models, $E_{\rm PISN} = 10^{51} \times [5.0 + 1.304
  (M_{\rm He} - 64)]$, where $M_{\rm He} = (13/24) \times (M_\star -
20) \Ms$ is the helium core (and equivalently the ejecta) mass and
$M_\star$ is the stellar mass.

@@

Here we present the gaseous and stellar evolution of two selected
halos in the simulation: one that has an early mass buildup but no
major mergers after $z=12$, and one that experiences a series of major
mergers between $z=10$ and $z=7$.  We name the halos ``quiet'' and
``intense'', respectively.  The entire simulation contains 38 galaxies
with 3640 Pop II stellar clusters and captures the formation of 333
Pop III stars.  The halo mass function and galaxy luminosities are
plotted in Figure \ref{fig:massfn}.

@@

Figure \ref{fig:sfr} shows the cosmic star formation rate (SFR) in the
simulation for Pop II and III stars.  Metal-free stars form at a
nearly constant rate of $5 \times 10^{-4}$ \sfr~until the end of the
simulation.  The dissociating and ionization radiation from nearby Pop
III stars and galaxies do not halt the formation of such stars in
low-mass haloes but only suppress them.  We expect a uniform
\hh~dissociating (Lyman-Werner) radiation background to further delay
them \citep{Machacek01, Wise07, OShea08}.  Metal-enriched star
formation begins early in the simulation at $z \sim 16$ when a PISN
blastwave overtakes a satellite halo with $M \sim 5 \times 10^5 \Ms$,
triggering star formation there.  It rises steadily as
$\rho_{\rm {\star,II}} \propto \exp(-z/1.1)$ until reaching $2 \times
10^{-2}$ \sfr~at $z=7$.  This value agrees with derived SFRs from
$z=6-10$ galaxy luminosity functions \citep[e.g.][]{Bouwens10} and
large-scale galaxy formation simulations \citep[e.g.][]{Schaye10}.

@@

Both halos start forming Pop II stars when $M = 10^7 \Ms$.  This is
consistent with the filtering mass $M_f$ of high-redshift halos when
it accretes mainly from a pre-heated IGM \citep{gnedin98, gnedin00,
  Wise08_Gal}.  Afterwards these halos can cool efficiently through
\hh~cooling, sustaining constant and sometimes bursting SF.  The
latter characteristics are equated with the definition of a galaxy.
The quiet halo forms $10^5 \Ms$ of stars by $z=9$.  This initial
starburst photo-evaporates the majority of its molecular clouds, in
addition to heating and ionizing the surrounding IGM out to a radius
of 10--15 kpc at $z=9$.  These respectively reduce the in-situ and
external cold gas supply that could feed future SF.

@@

J.~H.~W. is supported by NASA through Hubble Fellowship grant
\#120-6370 awarded by the Space Telescope Science Institute, which is
operated by the Association of Universities for Research in Astronomy,
Inc., for NASA, under contract NAS 5-26555.  M.~J.~T. and
M.~L.~N. acknowledge partial support from NASA ATFP grant NNX08AH26G.
Computational resources were provided by NASA/NCCS award SMD-09-1439.
J.~H.~W. thanks Renyue Cen, Amina Helmi, Marco Spaans, and Eline
Tolstoy for enlightening discussions.  The majority of the analysis
and plots were done with \texttt{yt} \citep{yt_full_paper}.

>>>>

\author{John H. Wise\altaffilmark{1,2}, 
  Matthew J. Turk\altaffilmark{3},
  Michael L. Norman\altaffilmark{4},
  Tom Abel\altaffilmark{5,6,7}}

\altaffiltext{1}{Department of Astrophysical Sciences, Princeton
  University, Peyton Hall, Ivy Lane, Princeton, NJ 08544}
\altaffiltext{2}{Hubble Fellow}
\altaffiltext{3}{Department of Astronomy, Columbia University, 538
  West 120th Street, New York, NY 10027}
\altaffiltext{4}{Center for Astrophysics and Space Sciences,
  University of California at San Diego, La Jolla, CA 92093}
\altaffiltext{5}{Kavli Institute for Particle Astrophysics and
  Cosmology, Stanford University, Menlo Park, CA 94025}
\altaffiltext{6}{Zentrum f\"ur Astronomie der Universit\"at, Institut
  f\"ur Theoretische Astrophysik, Albert-Ueberle-Str. 2, 69120
  Heidelberg, Germany}
\altaffiltext{7}{Heidelberg Institut f\"ur Theoretische Studien,
  Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany}

\email{jwise@astro.princeton.edu}

@@

We use mass-dependent luminosities and lifetimes of the Pop III stars
from \citet{Schaerer02}.  The radiation field is evolved with adaptive
ray tracing \citep{Abel02_RT, Wise11_Moray} and is coupled
self-consistently to the hydrodynamics.  We model the \hh~dissociating
radiation with an optically-thin, inverse square profile, centered on
all star particles.  These stars die as pair-instability SNe (PISNe)
if they are in the mass range 140--260 \Ms~\citep{Heger03}.  We use
the explosion energy from \citet{Heger02}, where we fit the following
function to their models, $E_{\rm PISN} = 10^{51} \times [5.0 + 1.304
  (M_{\rm He} - 64)]$, where $M_{\rm He} = (13/24) \times (M_\star -
20) \Ms$ is the helium core (and equivalently the ejecta) mass and
$M_\star$ is the stellar mass.

@@

\begin{figure}
\epsscale{1.15}
\plotone{z3}
\epsscale{1}
\caption{\label{fig:z3} Average halo metallicity from Pop III SNe as a
  function of total halo mass.  Inhomogenous metal enrichment causes
  the large spread in metallicity at low masses.  At $M_{\rm halo} > 3
  \times 10^7 \Ms$, all halos are enriched by Pop III SNe above [Z/H]
  $>$ --4.  The y-axis error bars show the 25\% and 75\% percentile
  values because the underlying distribution is not Gaussian.}
\end{figure}

Here we first present the global star formation rate (SFR) and metal
enrichment during early galaxy formation and then focus on the gaseous
and stellar evolution of two selected halos in the simulation: one
that has an early mass buildup but no major mergers after $z=12$, and
one that experiences a series of major mergers between $z=10$ and
$z=7$.  We name the halos ``quiet'' and ``intense'', respectively.
The entire simulation contains 38 galaxies with 3640 Pop II stellar
clusters and captures the formation of 333 Pop III stars.  The halo
mass function and galaxy luminosities are plotted in Figure
\ref{fig:massfn}.

@@

Figure \ref{fig:sfr} shows the cosmic SFR in the simulation for Pop II
and III stars.  Metal-free stars form at a nearly constant rate of $5
\times 10^{-4}$ \sfr~until the end of the simulation.  The
dissociating and ionization radiation from nearby Pop III stars and
galaxies do not halt the formation of such stars in low-mass haloes
but only suppress them.  We expect a uniform \hh~dissociating
(Lyman-Werner) radiation background to further delay them
\citep{Machacek01, Wise07, OShea08}.  Metal-enriched star formation
begins early in the simulation at $z \sim 16$ when a PISN blastwave
overtakes a satellite halo with $M \sim 5 \times 10^5 \Ms$, triggering
star formation there.  After $z=12$ in our simulation, it rises
steadily as $\rho_{\rm {\star,II}} \propto \exp(-z/1.1)$.  Star
formation commences gradually in the smallest galaxies because of
lowered gas fractions, causing a much steeper time dependency than
$\exp(-z/3)$ relation found in \citet{Hernquist03}.  Note that they
also acknowledged a steeper relation should exist at early times.  At
redshift 7, it reaches $2 \times 10^{-2}$ \sfr.  This value agrees
with derived SFRs from $z=6-10$ galaxy luminosity functions
\citep[e.g.][]{Bouwens10} and large-scale galaxy formation simulations
\citep[e.g.][]{Schaye10}.

\subsection{Pre-enrichment from Population III stars}

@@

Both halos start forming Pop II stars when $M = 10^7 \Ms$.  This is
consistent with the filtering mass $M_f$, which is plotted as a thick
line in Figure \ref{fig:evo}a, of high-redshift halos when it accretes
mainly from a pre-heated IGM \citep{gnedin98, gnedin00, Wise08_Gal}.
Afterwards these halos can cool efficiently through \hh~cooling,
sustaining constant and sometimes bursting SF.  The latter
characteristics are equated with the definition of a galaxy.  The
quiet halo forms $10^5 \Ms$ of stars by $z=9$.  This initial starburst
photo-evaporates the majority of its molecular clouds, in addition to
heating and ionizing the surrounding IGM out to a radius of 10--15 kpc
at $z=9$.  These respectively reduce the in-situ and external cold gas
supply that could feed future SF.

@@

J.~H.~W. thanks Renyue Cen, Amina Helmi, Marco Spaans, and Eline
Tolstoy for enlightening discussions.  J.~H.~W. is supported by NASA
through Hubble Fellowship grant \#120-6370 awarded by the Space
Telescope Science Institute, which is operated by the Association of
Universities for Research in Astronomy, Inc., for NASA, under contract
NAS 5-26555.  M.~J.~T. and M.~L.~N. acknowledge partial support from
NASA ATFP grant NNX08AH26G.  Computational resources were provided by
NASA/NCCS award SMD-09-1439.  TA acknowledges financial support from
the {\em Baden-W\"{u}rttemberg-Stiftung} under grant P-LS-SPII/18, the
Heidelberg Institut f\"ur Theoretische Studien. This work was
partially supported by NASA ATFP grant NNX08AH26G and NSF
AST-0807312. This research has made use of NASA’s Astrophysics Data
System Bibliographic Services.  The majority of the analysis and plots
were done with \texttt{yt} \citep{yt_full_paper}.
